Exploring probability distributions for bivariate temporal granularities

### 2019-12-09 main.Rmd
visualization intro update

Density estimates are useful in exploring the properties of the measured variable. They can give valuable indication of features like skewness, multimodality or tail behavior. The measured variable in the considered data structure can be dicrete or continuous. The probability distribution of them could be studied by estimating the probability density function (continuous) or probability mass function (dicrete). The approach can be parametric or non-parametric. A non-parametric approach makes less rigid assumptions about the distribution of the observed data making the data speak more for itself. These resonate more with the notion of EDA which advocates exploring data for patterns and relationships without requiring prior hypotheses. The entire distribution could be estimated using different density estimation methods or summarized using empirical quantiles. Several non-parametric density estimators and quantile estimators are proposed in the literature, the most commonly used form are as follows:

@@

Creating a visualisation requires a number of nuanced judgements. We exploit some features of layers and faceting [@Wilkinson1999-nk] and [@Wickham2009-pk] to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate temporal granularities. The general framework of visualizing the distribution invloves a faceting approach with one temporal granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. Different distribution plots might be appropriate depending on which features of the distrbution we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization.

@@

There are several ways to plot statistical distributions. The displayed probabilities in these plots are either computed using kernel density estimation methods or empirical statistical methods. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus. The probabilities displayed are based on actual data, which is aligned to principles that governed Tukey’s original boxplot. Density plots which uses a kernel density estimate to show the probability density function of the variable can show the entire distribution, unlike discretizing the distribution and showing only parts of it. They can be useful in spotting multimodality, however, they might become obscuring with too many categories and information on the entire distribution can add to cognitive load. Also, the probabilities in density plot are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations. The density based visualizations are subject to the same sample size restrictions and challenges that apply to any density estimation. In practice, the densities are estimated reasonably with at least 30 observations. Even with sample sizes of several hundred, however, choosing too large a value for bandwidth can cause the estimate to oversmooth the data.

@@

@@

Most commonly used techniques to display a distribution of data include the histogram (Karl Pearson), which shows the prevalence of values grouped into bins and the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary, displaying less detail than a histogram. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of these variations are helpful to get an idea of the distribution at a glance. Moreover, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

>>>>

Estimating the distribution of the measured variable can give valuable indication of features like central tendency, skewness, multimodality, tail behavior or variation. The measured variable in the considered data structure can be dicrete or continuous. The probability distribution of them could be studied by estimating the probability density function (continuous) or probability mass function (dicrete). The approach can be parametric or non-parametric. A non-parametric approach makes less rigid assumptions about the distribution of the observed data making the data speak more for itself. These resonate more with the notion of EDA which advocates exploring data for patterns and relationships without requiring prior hypotheses. The entire distribution could be estimated using different density estimation methods or summarized using empirical quantiles. Severalforms of  non-parametric density estimators and quantile estimators exist in the literature, the most commonly used form ( @Silverman1986-cl,  @Jones1992-pj) are as follows:

@@

Creating a visualisation requires a number of nuanced judgements. We exploit some features of layers and faceting [@Wilkinson1999-nk] and [@Wickham2009-pk] to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate cyclic temporal granularities with the data structure <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$. The framework involves faceting approach with one cyclic granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. Different distribution plots might be appropriate depending on which features of the distrbution we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization.

@@

There are several ways to plot statistical distributions. The displayed probabilities in these plots are either computed using kernel density estimation methods or quantile estimation methods. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus.  Density plots which uses a kernel density estimate to show the probability density function of the variable can show the entire distribution, unlike discretizing the distribution and showing only parts of it. The methods are subject to the same sample size restrictions and challenges that apply to any density estimation. In practice, the densities are estimated reasonably with at least 30 observations. Even with sample sizes of several hundred, however, choosing too large a value for bandwidth can cause the estimate to oversmooth the data.

@@

<!-- The probabilities displayed are based on actual data, which is aligned to principles that governed Tukey’s original boxplot. They can be useful in spotting multimodality, however, they might become obscuring with too many categories and information on the entire distribution can add to cognitive load. Also, the probabilities in density plot are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations. -->

@@

Most commonly used techniques to display a distribution of data include the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary, displaying less detail than a histogram. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of these variations are helpful to get an idea of the distribution at a glance. Moreover, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.
